# 6 Longtitudinal data

This week we work with two dataset, BPRS and RATS, and conduct the analyses in the MABS textbook chapter 8 and 9, but swapping the datasets..

## RATS

Dataset includes 16 rats, divided into 3 groups based on different diets.

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
ratsl <- read.csv(file = "~/IODS-project/data/ratsl.csv", row.names = 1)
# categorials into factors
ratsl$Group <- factor(ratsl$Group)
ratsl$ID <- factor(ratsl$ID)
# stardardize bprsl
ratsl <- ratsl %>% group_by(Time) %>% mutate(stweight = (Weight - mean(Weight)) / sd(Weight)) %>% ungroup()
str(ratsl)
```

## Graphical exploration of longtitudinal data

```{r}
# plot the individuals in time with standardized bprs value 
ggplot(ratsl, aes(x = Time, y = Weight, linetype = ID)) + geom_line() + facet_grid(. ~ Group, labeller = label_both) + scale_linetype_manual(values = rep(1:6, times=4)) + theme(legend.position = "none") +   scale_y_continuous(limits = c(min(ratsl$Weight), max(ratsl$Weight)))

```

All rats seem to gain weight. Group1 rats are lighter at the baseline, and in the end.
higher). In group 1, the between-subjects variability seems to decrease.

```{r}
# Summary plot
# Number of timepoints, baseline included
n <- ratsl$Time %>% unique() %>% length()
# Summarize mean and std error
ratssum <- ratsl %>% group_by(Group, Time) %>% summarise(mean = mean(Weight), se = ( sd(Weight) / sqrt(n) )) %>% ungroup()
glimpse(ratssum)

# plot the plots
par(mfrow = c(1,2))
# with means only
ggplot(ratssum, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
# added with the error bar
ggplot(ratssum, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```
Group means increase in time. Second plot includes the standard error bars.


```{r}
# boxplot
ggplot(ratsl, aes(x = factor(Time), y = Weight, fill = Group)) + geom_boxplot()

```

Boxplots with non-summarized data.

## Preparinf the summary measure approach

```{r}
# Summary measure approach, filter out outliers
# create the summary dataset with means from the weeks 1-8
rat8s <- ratsl %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
glimpse(rat8s)
ggplot(rat8s, aes(x=Group, y = mean)) + geom_boxplot() + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
```
There seems to be one outlier, let's get rid of that.

```{r}
# filter the outlier(s) and plot again
glimpse(rat8s)
rat8s1 <- filter(rat8s, Group == 2 & mean < 550 | Group == 1 & mean > 260 | Group == 3 & mean > 500)
glimpse(rat8s1)

ggplot(rat8s1, aes(x=Group, y = mean)) + geom_boxplot() + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")

```
Three outliers dropped, one from each group.



## Summary approach, comparing means

```{r}
summary(aov(mean ~ Group, data = rat8s1))

```

T-test to the summary data without the outliers, shows no difference between the groups (p=.60, conf.interval includes zero).


```{r}
kruskal.test(mean ~ Group, data = rat8s1)
```

Differences between group means are statistically significant according to both tests.

```{r}
# covariance analysis with the baseline
# add the baseline from the original wide dataset to the summary data
rats <- read.csv(file = "~/IODS-project/data/rats.csv", row.names = 1)
rat8s2 <- rat8s %>% mutate(baseline = rats$WD1)
glimpse(rat8s2)
# ANCOVA
fit <- lm(mean ~ baseline + Group, data = rat8s2)
summary(fit)
```

The covariance analysis reveals, that the baseline is highly related to the means in later time.
And, after controlling the baseline, there is no significant diffence left between the group means.

## BPRS

Do the growth profiles of the treatment groups differ?

```{r}
# read the data, in long form 
bprsl <- read.csv(file = "~/IODS-project/data/bprsl.csv", row.names = 1)
bprsl$treatment <- factor(bprsl$treatment)
bprsl$subject <- factor(bprsl$subject)
glimpse(bprsl)
```

```{r}
ggplot(bprsl, aes(x = weeks, y = bprs, group = subject)) + geom_line() + facet_grid(. ~ treatment, labeller = label_both) 
```
We have the individuals, all with bprs points decreasing in time, some more than others.


```{r}
bps_reg <- lm(bprs ~ weeks + treatment, data=bprsl)
summary(bps_reg)
```

Linear regression model with bprs as response and weeks and treatment group as explanatory variables.
From coefficient we see, that bprs decreases in time, but treatment group is not significant. 

### Random intercept model
```{r}
library(lme4)
bps_ref <- lmer(bprs ~ weeks + treatment + (1|subject), data = bprsl, REML=FALSE)
summary(bps_ref)

```

Note that in this model the within-subject dependencies are ignored. See std.dev of subject.

### Random intercept and random slope model 

```{r}
bps_ref2 <- lmer(bprs ~ weeks + treatment + (weeks|subject), data = bprsl, REML=FALSE)
summary(bps_ref2)
```

```{r}
anova(bps_ref, bps_ref2)
```

The model fit is better in model 2.

### Random intercept and random slope model with interaction

```{r}
bps_ref3 <- lmer(bprs ~ weeks * treatment + (weeks|subject), data = bprsl, REML=FALSE)
summary(bps_ref3)
```

The estimated regression parameters for the interaction  indicate that the bprs decrease slopes are somewhat higher for individuals in treatment group 2 than in group 1 (on average 0.71 higher). 


```{r}
anova(bps_ref2, bps_ref3)
```

The fit is not significantly better in model 3.


```{r}
# Create a vector of the fitted values
Fitted <- fitted(bps_ref2)
bprsl <- bprsl %>% mutate(Fitted)

# draw the plot of BPRSL with the observed bprs values
p1 <- ggplot(bprsl, aes(x = weeks, y = bprs, group = subject)) + geom_line() + facet_grid(. ~ treatment, labeller = label_both) + scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +  ylim(0,100) + theme(legend.position = "top") + ggtitle("Observed")

# draw the plot with fitted
p2 <- ggplot(bprsl, aes(x = weeks, y = Fitted, group = subject)) +
  geom_line() + facet_grid(. ~ treatment, labeller = label_both) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  ylim(0, 100) +
  theme(legend.position = "top") + ggtitle("Fitted")

p1; p2

```

